---
title: 'Clustering Customers for Segmentation'
author: "Chase Baggett"
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
  pdf_document: default
---
#Introduction
```{r,include=FALSE}
library(DT)
library(readxl)
library(GGally)
library(ggplot2)
library(networkD3)
# you may need to install the 'readxl' package first
eretail = read_excel("Online Retail.xlsx")
dim(eretail)
names(eretail)

eretail = eretail[eretail$Country != "Unspecified", ] # remove 'unspecified' country
eretail = eretail[eretail$Quantity > 0, ]             # remove returns/cancellations

IDtab = table(eretail$Country, eretail$CustomerID)   # crosstab country by customer ID
IDtab = apply(IDtab > 0, 2, sum)                      # is any customer ID duplicated across countries?
duplicateIDs = names(IDtab[IDtab > 1])               # duplicate IDs to clean up
eretail = eretail[!is.element(eretail$CustomerID, duplicateIDs), ]
rm(IDtab)

eretail$InvoiceMth = substr(eretail$InvoiceDate, 1, 7)         # extract month of invoice
eretail = eretail[as.vector(eretail$InvoiceMth) != "2011-12", ] # remove December 2011 as it only covers first week

eretail$Amount = eretail$Quantity * eretail$UnitPrice           # compute amount per invoice item
library(dplyr)
```
I am going to be using clustering techniques to attempt customer segmentation on a dataset of purchasing data for an online retailer. Specifically, I am analyzing customer spending patterns and habits. 

#Data

Our data is actually per product. Since I am not very concerned at the moment with what people are buying, but rather what types of customers they are in terms of regularity and spend, I'm going to aggregate the data to the invoice level which greatly reduces the size of our data. 

```{r,warning=FALSE,message=FALSE,fig.width=10}
data <-
  summarise(
  group_by(eretail, CustomerID, Country),
  Quantity = sum(Quantity),
  Unique_Items = n_distinct(StockCode),
  Spend = sum(UnitPrice * Quantity),
  Transactions = n_distinct(InvoiceNo),
  Days_Age = as.numeric(difftime(max(InvoiceDate),min(InvoiceDate),units="days"))
  )
data <- data[complete.cases(data),]
```

I am going to prep a few different variables on this dataset.

Field       | Description
------------|----------------------------------------
Spend       | Total Money Spent
Transactions| Number of Invoices
Quantity    | Number of Products Purchased
Unique_Items| Number of Unique Items Purchased
Days_Age    | Number of Days From First to Last Purchase

#Analysis

A basic summary of our variables makes it apparent that there are a large number of customers who placed a very small numbers of orders, had a very low age, and few transactions. In fact, 35% of customers only had one transaction. 


```{r,warning=FALSE,message=FALSE,fig.width=10}
pl_dat <- data[,c("Quantity","Transactions","Spend","Unique_Items","Days_Age")]
summary(pl_dat)
```

Viewing the remaining 65%, I'm going to log transform and scale all variables except Days_Age, which I will only scale. We have massively right tailed data and I am hoping that by using a log transformation the clusters we end up with will be logical. 

```{r,warning=FALSE,message=FALSE,fig.width=10}
log_zero <- function(x){ifelse(x==0,0,log(x))}
  
pl_dat$Quantity <- log_zero(pl_dat$Quantity)
pl_dat$Transactions <- log_zero(pl_dat$Transactions)
pl_dat$Spend <- log_zero(pl_dat$Spend)
pl_dat$Unique_Items <- log_zero(pl_dat$Unique_Items)
pl_dat <- as.data.frame(scale(pl_dat))
ggpairs(subset(pl_dat,Transactions>log(1)))
```

##Hierarchical Clustering

First I perform a basic hierarchical clustering and look at the dendrogram to identify a cut point. 

```{r,warning=FALSE,message=FALSE,fig.width=10}
hc <- hclust(dist(pl_dat))
plot(hc,labels=FALSE)
```

Now I am going to cut the tree down to 5 clusters and store the classification. 

We can take a look at a pairs plot of the scaled variables colored by classification.

```{r,warning=FALSE,message=FALSE,fig.width=10}
pl_dat$hclust <- as.factor(cutree(hc,5))
ggpairs(pl_dat,mapping=aes(color=hclust))
```


##K-Means Clustering

We will now use K-Means as well, using the 5 cluster we indentified using hierarchical clustering. 

Below, we see the results on the K-Means on our scaled data. 

```{r,warning=FALSE,message=FALSE,fig.width=10}
pl_dat <- data[,c("Quantity","Transactions","Spend","Unique_Items","Days_Age")]


log_zero <- function(x){ifelse(x==0,0,log(x))}

pl_dat$Quantity <- log_zero(pl_dat$Quantity)
pl_dat$Transactions <- log_zero(pl_dat$Transactions)
pl_dat$Spend <- log_zero(pl_dat$Spend)
pl_dat$Unique_Items <- log_zero(pl_dat$Unique_Items)
pl_dat <- as.data.frame(scale(pl_dat))

set.seed(1)
km <- kmeans(pl_dat,centers = 5)
pl_dat$kmeans <- as.factor(km$cluster)
ggpairs(pl_dat,mapping=aes(color=kmeans))
```

#Conclusion

We can see that we have some significant differences between k-means and hierarchical clustering. Most notably, kmeans has put the data into much closer to evenly sized clusters, whereas hclust, at the cut point I chose, places most data points into two clusters. 
```{r,warning=FALSE,message=FALSE,fig.width=10}
pl_dat$hclust <- cutree(hc,5)
res <- summarise(group_by(pl_dat,hclust,kmeans),cnt=n())
res <- res[order(res$cnt,decreasing = T),]
ggplot(res,aes(y=hclust,x=kmeans,size=cnt)) + 
  geom_point() + 
  theme(legend.position = "none")
```

We are going to look at hclust cluster #2, which is broken into 4 kmeans clusters. We can see that it is broken apart in what appears to be a very spherical way. 
```{r,warning=FALSE,message=FALSE,fig.width=10}
pl_dat$Combo <- paste(pl_dat$hclust,pl_dat$kmeans,sep="-")
ggpairs(subset(pl_dat,hclust==2),mapping=aes(color=Combo))
```

I am going to use the kmeans clustering from here on out. Using hclust to determine the number of clusters but using the underlying machinery of K-Means. 

Here, we can see the results of the clustering on the unscaled variables. 

```{r,warning=FALSE,message=FALSE,fig.width=10}
final_dat <-
  data[, c("Quantity",
  "Transactions",
  "Spend",
  "Unique_Items",
  "Days_Age")]
final_dat <- round(final_dat,2)
  final_dat$hclust <- as.factor(pl_dat$hclust)
  final_dat$kmeans <- as.factor(pl_dat$kmeans)
  
ggpairs(pl_dat,mapping=aes(color=kmeans))
```

With k means, the most desirable cluster for a customer to have is cluster 5, because is the highest spending, highest volume cluster. 

```{r,warning=FALSE,message=FALSE,fig.width=10}


  datatable(
  summarise(
  group_by(final_dat, kmeans),
  Transactions = round(mean(Transactions),2),
  Spend = round(mean(Spend),2),
  Unique_Items = round(mean(Unique_Items),2),
  Days_Age = round(mean(Days_Age),2)
  )
  )
```
We can then show which countries comprise the underlying clusters. 

```{r,warning=FALSE,message=FALSE,fig.width=8}
sankey_data <- pl_dat
sankey_data$Country <- data$Country
#sankey_data <- subset(sankey_data,Country!="United Kingdom")
top_countries <- summarise(group_by(sankey_data,Country),Cnt=n())
top_countries <- top_countries[order(top_countries$Cnt,decreasing = T),]$Country
sankey_data <- subset(sankey_data,Country %in% head(top_countries,25))
sankey_data$H1 <- paste("hclust",sankey_data$hclust)
sankey_data$H2 <- paste("kmeans",sankey_data$kmeans)
links1 <- sankey_data[,c("H2","Country")]
#links2 <- sankey_data[,c("Country","H2")]
colnames(links1) <- c("From","To")
#colnames(links2) <- c("From","To")
#links <- rbind(links1,links2)
links <- links1
nodes <- data.frame(node=unique(c(as.character(links$From),as.character(links$To))))
nodes$ID <- 1:nrow(nodes) - 1
colnames(nodes) <- c("From","Source")
links <- merge(nodes,links,by="From")
colnames(nodes) <- c("To","Target")
links <- merge(nodes,links,by="To")
links$value <- 1
nodes$name <- nodes$To

links <- summarise(group_by(links,Source,Target),value=sum(value))
sankeyNetwork(Links = links, Nodes = nodes, Source = "Source",
              Target = "Target", Value = "value", NodeID = "name",
              fontSize = 12, nodeWidth = 30)

```